---
title: "SpaDES 4 Dummies"
author: "Ceres Barros"
date: "November 27, 2017"
output: html_document
# output: pdf_document
---

## a VERY simple example of **`SpaDES`** awesomeness

This guide will take you through how to start your own model using **`SpaDES`**. It assumes you have already installed **`SpaDES`** and that it is working in your machine. If you haven't done this yet, please have a look at [this gist](https://gist.github.com/tati-micheletti/1de7505cdd387fe997f127f13eeb4393) to bypass known issues.



### The example
Let's imagine we want to understand whether the abundance of a species is statistically related with temperature. Both the abundance data and the temperature data are being constantly updated. Also, we want to have the possibility of analysing the relationship between the two iteratively, without needing to constantly change our script to account for the new data inputs.
In this example the abundance and temperature data are being updated by a simulation model.


### Before...
If we use R to develop our species abundance and temperature simulation models in the 'conventional way', we'll probably have a script were everything happens - the simulations and data analysis. At most, we would have a main script that sources others that contain useful functions, or are doing the simulations and data treatment/analysis separately. If you already use R like this, you'll find that the **`SpaDES`** way of thinking is similar. So why use it? Because it makes changing, adapting and sharing your code - or modules - much easier. 

Still don't believe me? Okay, let's solve our problem using the conventional way:

```{r the_r_way}
suppressMessages(library(raster))

## Template raster
r <- raster(nrows = 100, ncols = 100, xmn = -50, xmx = 50, ymn = -50, ymx = 50)

## SPECIES ABUNDANCE "SIMULATIONS"
## Our VERY simple "simulation" model merely generates rasters that follows a Gaussian distribution
abundance_model <- function(r, Time) {
  abund_outputs <- list()
  for(t in 1:Time) { 
    abund_outputs[[t]] <- SpaDES.tools::gaussMap(r, scale = 100, var = 0.03) 
  }
  return(abund_outputs)
}

## Lenght of the simulation (or simply the number of model iterations)
Time <- 10

## Run the model nad plot results
abundance <- abundance_model(r = r, Time = Time)
plot(stack(abundance))

## TEMPERATURE "SIMULATIONS"
## The temperature simulation model will be almost the same as the vegetation one - remember this is a dummy example ;)
r <- raster(nrows = 100, ncols = 100, xmn = -50, xmx = 50, ymn = -50, ymx = 50)
temp_model <- function(r, Time) {
  temp_outputs <- list()
  for(t in 1:Time) { 
    temp_outputs[[t]] <- SpaDES.tools::gaussMap(r, scale = 100, var = 0.1) 
  }
  return(temp_outputs)
}

## Lenght of the simulation (or simply the number of model iterations)
Time <- 10

## Run the model
temperature <- temp_model(r = r, Time = Time)
plot(stack(temperature))

## DATA ANALYSIS
stats_analysis <- function(Data){
  lm1 <- lm(abund ~ temp, data = Data)
  plot(Data$abund ~ Data$temp, xlab = "Temp.", ylab = "Species abundance")
  abline(a = lm1$coefficients["(Intercept)"], b = lm1$coefficients["temp"], lwd = 2, col = "blue")
}

par_sets <- par(mfrow = c(2, 5))
for(t in 1:Time){
  outputdata <- data.frame(abund = abundance[[t]][], temp = temperature[[t]][])
  stats_analysis(Data = outputdata)
}
```

```{r, echo = FALSE}
par(par_sets)
```


### After...
Let's now solve the same problem using the **`SpaDES`** approach. I like to start by creating a *global.R* script (you can call it whatever you want, though) that will load the **`SpaDES`** libraries and run **`SpaDES`** simulations.

```{r globalscript}
suppressMessages(library(SpaDES))  ## should automatically download all packages in the SpaDES family and their dependencies

## decide where you're working
setPaths(cachePath = "cache",
         inputPath = "inputs",
         modulePath = "m",
         outputPath = "outputs")
getPaths() ## check that this is what you wanted

## Let's create a self-contained module that will simulate the species' abundance for any given period of time and frequency.
if(!dir.exists(file.path(getPaths()$modulePath, "speciesAbundance"))){
  newModule(name = "speciesAbundance", path = getPaths()$modulePath)
}

```

You will notice that `newModule` has created a module folder inside */m* that contains both the module *.R* script template, as well as the documentation template (the *.Rmd* file). Although we will not be discussing the *.Rmd* file, please bear in mind that this is a **fundamental** part of creating a reproducible and transparent module. The documentation should contain not only the description of the module, but also some simple examples of what it can do.

`newModule` also created the folder */data* where data necessary to the module should be put in, and the folder */tests* that may contain testing scripts. We won't be using neither of them in this example. 

**/!\\ Attention /!\\** : `newModule` should only be run once, or separately in the console, otherwise you may lose your module edits by re-running it again - this is why I wrapped it in an `if` statement (smart eh?).


Now go ahead and open the *speciesAbundance.R* script. I know, I know... It seems confusing and there's a lot of stuff in that template! Let's break it down:

The first "bit" of code is basically defining the module metadata. It will allow you to define who the author is, keywords that describe your module and the module(s) and package(s) version(s). It will also define parameters and inputs that the module requires, and the outputs it produces.

This module requires no input data, so we leave that bit empty. As for the outputs, it produces a template raster (produced during the `abundanceInit` event) and a list of abundance rasters (produced during the `abundanceSim` event). So we define these two outputs in the funciton.

You'll notice that I have removed several parameters that we don't need (like save parameters).

If you are unsure of what input and output parameters are in the context of a module, a good rule of thumb is that inputs are all the `sim$...` objects that appear to the **right-hand side** of a `<-`, whereas output parameters are the `sim$...` objects appear to the **left-hand side** of a `<-`.


Here's how I defined the module:
```{r abund_module, eval=FALSE}
defineModule(sim, list(
  name = "speciesAbundance",
  description = "Species abundance simulator",
  keywords = c("species", "abundance", "gaussian", "spatial"),
  authors = person("Mr.", "Me", email = "mr.me@example.com", role = c("aut", "cre")),
  childModules = character(0),
  version = list(SpaDES.core = "0.1.0", speciesAbundance = "0.0.1", SpaDES.addins = "0.1.0", SpaDES.tools = "0.1.0"),
  # spatialExtent = raster::extent(rep(NA_real_, 4)),
  timeframe = as.POSIXlt(c(NA, NA)),
  timeunit = "year",
  citation = list("citation.bib"),
  documentation = list("README.txt", "speciesAbundance.Rmd"),
  reqdPkgs = list("raster"),
  parameters = rbind(
    # defineParameter("paramName", "paramClass", value, min, max, "parameter description"),
    defineParameter("simulationTimeStep", "numeric", 1, NA, NA, "This describes the simulation time step interval"),
    defineParameter(".plotInitialTime", "numeric", 1, NA, NA, "This describes the simulation time at which the first plot event should occur"),
    defineParameter(".plotInterval", "numeric", 1, NA, NA, "This describes the simulation time interval between plot events")
  ),
  inputObjects = bind_rows(
    #expectsInput("objectName", "objectClass", "input object description", sourceURL, ...),
  ),
  outputObjects = bind_rows(
    #createsOutput("objectName", "objectClass", "output object description", ...),
    createsOutput(objectName = "r", objectClass = "RasterLayer", desc = "Template raster"),
    createsOutput("abundRasters", "RasterLayer", "Raster layer of species abundance at any given year")
  )
))

```

Th rest of the script defines the event sequence for this module - remember **`SpaDES`** = Spatial Discrete Event Simulator - and the events themselves. 

The initialisation (`init`) event is **mandatory**, the others are not. **`SpaDES`** knows how to deal with an `init` event and "searches" for it. For example, it knows that the initialisations of all active modules will be executed first. 

The order in which modules (and consequently their initialisations) are executed is automatically determined by inter-module dependencies (i.e. module inputs that are the outputs of other modules) - I told you **`SpaDES`** is cool! If there are no inter-module dependencies the order is determined by the other in which modules are listed in the *global.R* script **and/or** by the way we schedule the events. But we'll get to that part later.

Since we are only interested in simulating and plotting species abundances, we can remove unnecessary events from the script. I have only kept 3 events: the initialisation, aplotting event and an abundance simulation event. 

Let's look at the **event functions** first (those that come after `doEvent.speciesAbundance`). I have renamed the event functions, and `Init` became `abundanceInit`, `Plot` became `abundancePlot` and `event1` became `abundanceSim`.


####Initialisation function

In the initialisation, we are creating a template raster and a storage list for our species abundance outputs (which will also be rasters). Notice how the only argument to `abundanceInit` is the `sim` object, which is where the objects being created are stored. Always take care to `return()` the `sim` object at the end of an event.
```{r abund_module2, eval = FALSE}
abundanceInit <- function(sim) {
  ## Template raster
  sim$r <- raster(nrows = 100, ncols = 100, xmn = -50, xmx = 50, ymn = -50, ymx = 50)
  
  ## create storage list of species abundance
  sim$abundRasters <- list()
  
  return(invisible(sim))
}
```



####Abundance simulation event function

This event is basically the "prima donna"" of this module. This is where we will generate species abundances. Notice how instead of a loop, we now have the event running the `abundance_model` function (which we'll define separately below) and storing its outputs in the `sim$abundRaster` object. Notice as well that we use `time(sim)` as the identifier of the list slots where outputs are stored. Go ahead and check what `SpaDES.core::time` is doing!
Again, the sole argument and output to this event function is the `sim` object - notice a pattern? ;)
```{r abund_module3, eval = FALSE}
abundanceSim <- function(sim) {
  ## Generate species abundances - our "simulation"
  sim$abundRasters[[time(sim)]] <- abundance_model(ras = sim$r)
  
  return(invisible(sim))
}
```

####Plotting event function:

Yup, you got it. This event simply plots the species abundance rasters that are produced during the `abundanceSim` event. "How does it know that it should plot them **after** the abundance generator?" you ask me puzzled. Good question sir/madam, 5 points for Griffindor!
The order of events is defined as a `doEvent.<module_name>` function - whose template is automatically created under the `defineModule` function. Shall we have a look?
```{r abund_module4, eval = FALSE}
abundancePlot <- function(sim) {
  ## plot abundances
  plot(sim$abundRasters[[time(sim)]], 
       main = paste0("Species abundance\nat time ", time(sim)))
  
  return(invisible(sim))
}
```


####Scheduling events

Now that we've defined our evetn functions, we need to set the order in which they will happen for this module. For example, we don't want the plotting to occur before the generation of species abundances, so we'll schedule all `abundancePlot` events to occur slightly after the   `abundanceSim` events.

Events are scheduled using the `doEvent.<module_name>` function: 
```{r abund_module5, eval = FALSE}
doEvent.speciesAbundance = function(sim, eventTime, eventType, debug = FALSE) {
  switch(
    eventType,
    init = {
      ## do stuff for this event
      sim <- abundanceInit(sim)
      
      ## schedule future event(s)
      sim <- scheduleEvent(sim, eventTime = start(sim), moduleName = "speciesAbundance", eventType = "SimulAbund")
      sim <- scheduleEvent(sim, eventTime = P(sim)$.plotInitialTime, moduleName = "speciesAbundance", eventType = "plot")
    },
    plot = {
      ## do stuff for this event
      sim <- abundancePlot(sim)
      
      ## schedule future event(s)
      sim <- scheduleEvent(sim, eventTime = time(sim) + P(sim)$.plotInterval, moduleName = "speciesAbundance", eventType = "plot")
    },
    SimulAbund = {
      ## do stuff for this event
      sim <- abundanceSim(sim)
      
      ## schedule future event(s)
      sim <- scheduleEvent(sim, eventTime = time(sim) + P(sim)$simulationTimeStep, moduleName = "speciesAbundance", eventType = "SimulAbund")
    },
    warning(paste("Undefined event type: '", current(sim)[1, "eventType", with = FALSE],
                  "' in module '", current(sim)[1, "moduleName", with = FALSE], "'", sep = ""))
  )
  return(invisible(sim))
}
```
Now this one seems complicated... But it's not, trust me.

I encourage you to check what `base::switch` is doing on your own, but if you're lazy here's a dummy explanation: it tells R that the behaviour of your function will change (or switch) depending on the value of `eventType`. So we need to define the behaviour of `doEvent` for all the event types we have in this module. In each of them we will execute the event functions and schedule future events with `scheduleEvent`.

The first one is, obviously, `init` - don´t change the name of this one. 
In `init` we run the initialisation event and schedule the abundance simulation and plotting events. Notice that the `abundanceSim` event will be generated at `start(sim)` (i.e. at the starting time), which basically is the same as running it during the initialisation. The first plotting event will be executed at the time defined by the `.plotInitialTime` parameter, which is stored in the `sim` object (and obtained using  `SpaDES.core::P`). 

The `abundancePlot` plotting event type is defined next. 
Besides doing the obvious plotting, it **schedules itself** to occur at a frequency defined by the `.plotInterval` parameter - isn't that cool? I think it is, but I'm only a silly ecologist.

Finally, we define the `SimulAbund` event during which the species' abundances will be generated. 
It is similar to the plotting event, with an important difference being that it uses the `simulationTimeStep` parameter to re-schedule itself.

**Note:** Notice how I've scheduled future events to `time(sim) + P(sim)$simulationTimeStep` or `time(sim) + P(sim)$.plotInterval`.

#### Additional functions

Ok, just one more bit left. Events may rely on other funcitons that can either be sourced from other scripts, or defined at the end of the module script. 
This is the case for the species abundances generator function, which I have coded at the end of my *speciesAbundance.R* script:
```{r abund_module6, eval = FALSE}
abundance_model <- function(ras) {
  abund_ras <- SpaDES.tools::gaussMap(ras, scale = 100, var = 0.01) 
  return(abund_ras)
}
```

#### Creating additional modules

Now go ahead and repeat the previous steps to create a second module that will generate yearly temperatures.

Apart from changing the objects and functions names, I have also included the template raster `r` as an input object for the temperature module - remember that `r` is created during the `abundanceInit`. This avoids repeating the creation of the template raster.
This may not seem like a big deal in our example, but it can be if you're generating heavy objects, or relying on functions that take a while to run.

Here's how my final *temperature.R* script looks like:

```{r temp_module, eval = FALSE}
defineModule(sim, list(
  name = "temperature",
  description = "Temperature simulator",
  keywords = c("temperature", "gaussian", "spatial"),
  authors = person("Mr.", "Me", email = "mr.me@example.com", role = c("aut", "cre")),
  childModules = character(0),
  version = list(SpaDES.core = "0.1.0", speciesAbundance = "0.0.1", temperature = "0.0.1", SpaDES.addins = "0.1.0", SpaDES.tools = "0.1.0"),
  spatialExtent = raster::extent(rep(NA_real_, 4)),
  timeframe = as.POSIXlt(c(NA, NA)),
  timeunit = "year",
  citation = list("citation.bib"),
  documentation = list("README.txt", "temperature.Rmd"),
  reqdPkgs = list("raster"),
  parameters = rbind(
    #defineParameter("paramName", "paramClass", value, min, max, "parameter description"),
    defineParameter("simulationTimeStep", "numeric", 1, NA, NA, "This describes the simulation time step interval"),
    defineParameter(".plotInitialTime", "numeric", 1, NA, NA, "This describes the simulation time at which the first plot event should occur"),
    defineParameter(".plotInterval", "numeric", 1, NA, NA, "This describes the simulation time interval between plot events")
  ),
  inputObjects = bind_rows(
    #expectsInput("objectName", "objectClass", "input object description", sourceURL, ...),
    expectsInput("r", "RasterLayer", "Template raster")
  ),
  outputObjects = bind_rows(
    #createsOutput("objectName", "objectClass", "output object description", ...),
    createsOutput("tempRasters",  "list", "List of raster layers of temperature at any given year")
  )
))

## event types
#   - type `init` is required for initialiazation

doEvent.temperature = function(sim, eventTime, eventType, debug = FALSE) {
  switch(
    eventType,
    init = {
      ## do stuff for this event
      sim <- temperatureInit(sim)
      
      ## schedule future event(s)
      sim <- scheduleEvent(sim, eventTime = start(sim), moduleName = "temperature", eventType = "SimulTemp")
      sim <- scheduleEvent(sim, eventTime = P(sim)$.plotInitialTime, moduleName = "temperature", eventType = "plot")
    },
    plot = {
      ## do stuff for this event
      sim <- temperaturePlot(sim)
      
      ## schedule future event(s)
      sim <- scheduleEvent(sim, eventTime = time(sim) + P(sim)$.plotInterval, moduleName = "temperature", eventType = "plot")
    },
    SimulTemp = {
      ## do stuff for this event
      sim <- temperatureSim(sim)
      
      ## schedule future event(s)
      sim <- scheduleEvent(sim, eventTime = time(sim)+ P(sim)$simulationTimeStep, moduleName = "temperature", eventType = "SimulTemp")
    },
    warning(paste("Undefined event type: '", current(sim)[1, "eventType", with = FALSE],
                  "' in module '", current(sim)[1, "moduleName", with = FALSE], "'", sep = ""))
  )
  return(invisible(sim))
}

## This is the 'init' event:
temperatureInit <- function(sim) {
  ## create storage list of species temperature
  sim$tempRasters <- list()
  
  return(invisible(sim))
}

## This is the plotting event funciton
temperaturePlot <- function(sim) {
  ## plot temperature
  plot(sim$tempRasters[[time(sim)]], 
       main = paste0("Temperature\nat time ", time(sim)))
  
  return(invisible(sim))
}

## This is the temperature simulation event function
temperatureSim <- function(sim) {
  ## Generate temperature - our "updated data"
  sim$tempRasters[[time(sim)]] <- temperature_model(ras = sim$r)
  
  return(invisible(sim))
}

## This is not an event, but a function that we define separately 
## and that contains our "simulation model"
temperature_model <- function(ras) {
  temp_ras <- SpaDES.tools::gaussMap(ras, scale = 100, var = 0.01) 
  return(temp_ras)
}
```

#### Modules that depend on other modules

Our third and last module will be used to run the statistical analysis at each year, after the abundances and temperatures are generated.
Hence, it'll depend on the outputs of the `speciesAbundance` and the `temperature` modules. We'll call it `speciesTempLM`.

The interest of keeping the statistical analysis in a separate module lies on the fact that it allows us to easily swap and compare different statistical models to analyse our data if we want to - either by changing the module code, or by having other modules with different statistical models. 

It also allows for greater flexibility when it comes to **when** the statistical model is supposed to run. For example, imagine that instead of every year, we want to fit it at every 5 years, using the previous 5 years of data. By having the statistical analysis contained in its own module, we don't need to fiddle with the other module scripts in order to make these changes.

Finally, I'd like to draw your attention to a few differences in this module's script:

* The frequency of the statistical analysis (and correspondent plots) will be determined by the parameter `statsTimestep`. This parameter also determines the number of data years to be used to fit the linear model. If `statsTimestep` = 5, then the statistical analysis will use the precedent 5 years of data including the year in which the event is running (a total of 6 years of data);

* This module **requires inputs**. They are specified in `inputObjects` part of `defineModule` - notice how I've respected the names, classes and description of the objects that come from the `speciesAbundance` and the `temperature` modules;

* This time, we have **two additional functions** at the end of the script: the function fitting the linear model and a plotting function.

Here's the full module script.
Notice how the future events where scheduled to `P(sim)$statsTimestep + 0.1`. This forces the statistic analyses to occur **after** the abundance and temperature rasters are ready.
```{r stats_module, eval = FALSE}
# Everything in this file gets sourced during simInit, and all functions and objects
# are put into the simList.
defineModule(sim, list(
  name = "speciesTempLM",
  description = "Statistical analysis of species ~ temperature relationships using LM",
  keywords = c("linear model"),
  authors = person("Mr.", "Me", email = "mr.me@example.com", role = c("aut", "cre")),
  childModules = character(0),
  version = list(SpaDES.core = "0.1.0", speciesTempLM = "0.0.1", speciesAbundance = "0.0.1", temperature = "0.0.1", SpaDES.addins = "0.1.0", SpaDES.tools = "0.1.0"),
  spatialExtent = raster::extent(rep(NA_real_, 4)),
  timeframe = as.POSIXlt(c(NA, NA)),
  timeunit = "year",
  citation = list("citation.bib"),
  documentation = list("README.txt", "speciesTempLM.Rmd"),
  reqdPkgs = list("raster"),
  parameters = rbind(
    #defineParameter("paramName", "paramClass", value, min, max, "parameter description"),
    defineParameter("statsTimestep", "numeric", 1, NA, NA, "This describes the how often the statitiscal analysis will be done")
  ),
  inputObjects = bind_rows(
    #expectsInput("objectName", "objectClass", "input object description", sourceURL, ...),
    expectsInput( "abundRasters", "list", "List of raster layers of species abundance at any given year"),
    expectsInput( "tempRasters", "list", "List of raster layers of temperature at any given year")
  ),
  outputObjects = bind_rows(
    #createsOutput("objectName", "objectClass", "output object description", ...),
    createsOutput("outputdata", "list", "List of dataframes containing species abundances and temperature values per pixel"),
    createsOutput( "outputLM", "list", "List of output yearly LMs (abundance ~ temperature)"),
    createsOutput( "yrs", "numeric", "Vector of years used for statistical analysis")
  )
))

## event types
#   - type `init` is required for initialiazation

doEvent.speciesTempLM = function(sim, eventTime, eventType, debug = FALSE) {
  switch(
    eventType,
    init = {
      ## do stuff for this event
      sim <- statsInit(sim)

      ## schedule future event(s)
      sim <- scheduleEvent(sim, P(sim)$statsTimestep + 0.1, "speciesTempLM", "stats")
      sim <- scheduleEvent(sim, P(sim)$statsTimestep + 0.1, "speciesTempLM", "plot")
    },
    plot = {
      ## do stuff for this event
      sim <- statsPlot(sim)
      
      ## schedule future event(s)
      sim <- scheduleEvent(sim, time(sim) + P(sim)$statsTimestep, "speciesTempLM", "plot")
    },
    stats = {
      ## do stuff for this event
      sim <- statsAnalysis(sim)
      
      ## schedule future event(s)
      sim <- scheduleEvent(sim, time(sim) + P(sim)$statsTimestep, "speciesTempLM", "stats")
    },
    warning(paste("Undefined event type: '", current(sim)[1, "eventType", with = FALSE],
                  "' in module '", current(sim)[1, "moduleName", with = FALSE], "'", sep = ""))
  )
  return(invisible(sim))
}

## template initialization
statsInit <- function(sim) {
  ## create outputs storage lists
  sim$outputdata <- list()
  sim$outputLM <- list()
  
  return(invisible(sim))
}

## Plotting event
statsPlot <- function(sim) {

  plotLMResults(Data = sim$outputdata[[time(sim)]], model = sim$outputLM[[time(sim)]])
  
  return(invisible(sim))
}

## Statistical analysis event
statsAnalysis <- function(sim) {
  
  sim$yrs <- seq(time(sim) - P(sim)$statsTimestep + 1, time(sim), 1)

  sim$outputdata[[time(sim)]] <- do.call(rbind.data.frame, 
                                         lapply(sim$yrs, FUN = function(y){
                                           temp <- data.frame(abund = sim$abundRasters[[y]][], temp = sim$tempRasters[[y]][], year = y)          
                                           return(temp)
                                         }))
  
  sim$outputLM[[time(sim)]] <- linearModel(Data = sim$outputdata[[time(sim)]])
  
  return(invisible(sim))
}

## Other functions
linearModel <- function(Data){
  return(lm1 <- lm(abund ~ temp, data = Data))
}

plotLMResults <- function(Data, model){
  plot(Data$abund ~ Data$temp, xlab = "Temp.", ylab = "Species abundance", 
       main = paste("From years", min(Data$year)-0.1, "to", max(Data$year)-0.1, sep = " "))
  abline(a = model$coefficients["(Intercept)"], b = model$coefficients["temp"], lwd = 2, col = "blue")
}
```


### Let's play!
Ok, now that we have our modules ready we can set up the simulation. Let's go back to our *global.R* script. 

The function `simInit` will take a bunch of parameter lists and will prepare a simulation object that can later be run by the `spades` function. Let's see what these lists consist of:

The first list, `modules`, contains modules we want to activate.

Then, `times` is a list containing the start and end times of the simulation and what time units we're working with - so it defines the length of the simulation. It's important that the start and ending times are defined in decimals, because **`SpaDES`** allows decomposing time units into smaller fractions - and we'll take advantage of this feature!

`parameters` is a list of lists. It contains the values of parameters used by each modules, and well as "global" parameters used by all of them (`.globals`). Note that not all module parameters need to be defined in this way. In fact, a module can use a parameter that is listed inside the module *.R* script (and has a default value set there), but is never listed in `simInit`. An example of this would be `.plotInterval` which used and defined in the `speciesAbundance` and `temerature` modules. Conversely, `.plotInitialTime` is a parameter used and defined in these modules, but whose value we change when running `simInit`.
To make sure that plotting always occurs after the abundances are generated, I have changed `.plotInitialTime` to 1.5. Since `.plotInterval` remains unchanged (remember, that is its defined within the module scripts), all plots will occur at the "half" year (1.5, 2.5, 3.5, etc...) 
Finally, `paths` contains the directory paths that we set earlier.

```{r globalscript2}
## list the modules to use
modules <- list("speciesAbundance", "temperature", "speciesTempLM")

## Set simulation and module parameters
times <- list(start = 1.0, end = 10.1, timeunit = "year")
parameters <- list(
  .globals = list(simulationTimeStep = 1, .plotInitialTime = 1.5),
  speciesTempLM = list(statsTimestep = 5)
)

## make a list of directory paths
paths <- getPaths()

## Simulation setup
mySim <- simInit(times = times, params = parameters, 
                modules = modules, paths =  paths)
```


####Have you set up your simulation correctly?

Before starting the simulations we should check if the modules were linked correctly.

**Module diagram**

`moduleDiagram` is a useful function that shows module interdependencies as a network. The direction of the arrows indicates an output to input flow. You can see that the outputs of the `speciesAbundance` and `temperature` modules are inputs to the `speciesTempLM` module.
```{r modulediagram, echo = -1}
clearPlot()
moduleDiagram(mySim)
```

**Object diagram**

`objectDiagram` provides yet another way of checking if the modules are linked correctly, by explicitly showing the objects that pass between modules.
```{r eventdiagram, echo = -1}
clearPlot()
objectDiagram(mySim)
```

####Run **`SpaDES`**!

Ok, everything seems to be correct so far. Let's try to run the simulations. 
I have used `debug = TRUE` so that `spades` prints the events as they are being executed. In case something fails, we'll know where it stopped.

```{r spadesrun, eval = FALSE}
## run simulation
dev() # on Windows and Mac, this opens external device if using Rstudio, it is faster
clearPlot()
spades(mySim, debug = TRUE)
```



### Additional notes

* **`SpaDES`** is an extremely powerful package, whose potential goes well beyond what has been discussed in this dummy example. If you don't feel so dummy any more and want to explore it further, have a look at the [LCC2005]() and the [wolfALps]() module tutorials. Also, do go to the [**`SpaDES`** webpage](http://predictiveecology.org/) to find further information about the package, as well as upcoming workshops and publications.
